Kernel k-Groups via Hartigan's Method

### 2018-08-10 energy_clustering/pami18/energy_clustering.tex
fixed # of sigdigs

The following experimental setup holds unless specified
otherwise. 
We consider
the metrics
\begin{align}
\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, 
%\kk_{\alpha}(x,y) = \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
%- \| x-y \|^{\alpha} \right), 
\label{eq:rho_alpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  
%\widetilde{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|}{2\sigma}}, 
\label{eq:rho_tilde}\\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}},
%\widehat{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
\label{eq:rho_hat}
\end{align}
which define the 
corresponding kernels through 
\eqref{eq:kernel_semimetric}, where we always fix $x_0=0$.
%The standard metric from energy statistics is $\rho_1$ and 
%will always be present in the experiments as a reference.
We consider the weights associated to data points to be $w(x) = 1$ for
all $x$, so that $\W = I$ in 
Algorithms~\ref{kmeans_algo}~and~\ref{algo}.
For k-means, GMM and spectral clustering we use the
implementations of \emph{scikit-learn} library \cite{scikit-learn}, where  
k-means is initialized with 
k-means++ \cite{Vassilvitskii}, 
and GMM is initialized with the output of k-means, making 
it more robust and preventing
it from breaking in high dimensions. 
The spectral clustering implementation of scikit-learn 
is based on \cite{Malik}. 
Kernel k-means is implemented according to Algorithm~\ref{kmeans_algo} while
kernel k-groups follows Algorithm~\ref{algo}.
Both will also be initialized with k-means++, unless specified otherwise.
We run every algorithms $5$ times with different initializations and then
we choose the result with the best objective function value. 
We evaluate the clustering quality by
the accuracy defined as
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi
\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij} ,
\end{equation}
where $\hat{Z}$ is the predicted label matrix, $Z$
is the ground truth label matrix, 
and $\pi$ is a permutation of $\{1,2,\dotsc,k\}$. 
Thus, the accuracy corresponds to the fraction
of correctly classified data points, and it is always between $[0,1]$.
For each setting, we show the average accuracy over $100$ Monte
Carlo trials, with error bars indicating the standard error.

@@

We first consider one-dimensional data for a two-class problem. We compare
kernel k-groups with k-means and GMM, as illustrated in 
Fig.~\ref{fig:1dexperiment}. We have two mixtures of normal and 
lognormal distributions. Notice that in the kernel density estimation plots
for lognormal distribution, only kernel k-groups was able to distinguish
between the two classes. The accuracy results for both density estimation
cases are in Table~\ref{tb:dens}. We remark that kernel k-groups in
this one-dimensional example performed the same as the deterministic
Algorithm~\ref{algo1d} introduced in the Appendix.

\begin{table}[h]
\centering
\caption{
\label{tb:dens} Accuracy results for the density
estimation of Fig.~\ref{fig:1dexperiment}d--e.
}
\begin{tabular}{@{}l|ll@{}}
       Method        & normal & lognormal \\ \midrule[.5pt]
        kmeans       & 0.773  & 0.502     \\
         GMM         & 0.881  & 0.513     \\
   kernel k-groups   & 0.802  & 0.847  
\end{tabular}
\end{table}

@@

\begin{table}
\caption{
\label{tb:cigar_circle}
Clustering the data from Fig.~\ref{fig:2dscatter}.
For spcetral clustering, kernel k-means and kernel k-groups we use
the metric $\widetilde{\rho}_2$ (see \eqref{eq:rho_tilde}) for the data
in Fig.~\ref{fig:2dscatter}a, while $\widehat{\rho}_1$ (see \eqref{eq:rho_hat})
for the data in Fig.~\ref{fig:2dscatter}b. We have 800
points on each trial and 30 Monte Carlo runs for both datasets.
}
\centering
\begin{tabular}{@{}l|ll|ll@{}}
                   & \multicolumn{2}{c|}{Fig.~\ref{fig:2dscatter}a}  &
                     \multicolumn{2}{c}{Fig.~\ref{fig:2dscatter}b}  \\
       Method      & Accuracy & SEM     & Accuracy & SEM \\ \midrule[.5pt]
        kmeans       & 0.5330 & 0.0048  & 0.5210   & 0.0027 \\
         GMM         & 0.9292 & 0.0295  & 0.5330   & 0.0039 \\
 spectral clustering & 0.5775 & 0.0096  & 0.7253   & 0.0028 \\
    kernel k-means   & 0.9999 & 0.0002  & 0.9999   & $4.16\, 10^{-5}$ \\
   kernel k-groups   & 0.9987 & 0.0003  & 0.9999   & $4.16\, 10^{-5}$  
\end{tabular}
\end{table}

@@

\begin{table}
\caption{
\label{tb:dermatology_accuracy}
For the dataset \cite{Dua2017,Guvenir1998} (see also
Table~\ref{tb:dermatology}) we show the accuracy \eqref{eq:accuracy} and
the adjusted Rand index (aRand) of several methods.
In \cite{RizzoClustering} the authors 
obtained $\textnormal{aRand}=0.9195$ using an energy statistics
based method, while \cite{Kgroups} obtains $\textnormal{aRand}=0.9188$ 
where points with missing
entries are removed. Below we complete the missing entries with the mean.
If we remove the points with missing entries, kernel k-groups provides an
improvement of $\textnormal{accuracy}=0.9637$ and $\textnormal{aRand}=0.9396$.
}
\centering
\begin{tabular}{@{}l|ll@{}}
       Method      & Accuracy & aRand \\ \midrule[.5pt]
        kmeans       & 0.7131 & 0.6897 \\
         GMM         & 0.8770 & 0.8403 \\
 spectral clustering & 0.9536 & 0.9125 \\
    kernel k-means   & 0.7514 & 0.8510 \\
   kernel k-groups   & \textbf{0.9617} & \textbf{0.9356}
\end{tabular}
\end{table}

>>>>

%The following experimental setup holds unless specified otherwise. 
We consider
the metrics
\begin{align}
\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, 
%\kk_{\alpha}(x,y) = \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
%- \| x-y \|^{\alpha} \right), 
\label{eq:rho_alpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  
%\widetilde{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|}{2\sigma}}, 
\label{eq:rho_tilde}\\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}},
%\widehat{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
\label{eq:rho_hat}
\end{align}
which define the 
corresponding kernels through 
\eqref{eq:kernel_semimetric}, where we always fix $x_0=0$.
We use $\rho_\alpha$ by default, unless otherwise specified.
%will always be present in the experiments as a reference.
We consider the weights associated to data points to be $w(x) = 1$ for
all $x$, so that $\W = I$ in 
Algorithms~\ref{kmeans_algo}~and~\ref{algo}.
For k-means, GMM and spectral clustering we use the
implementations of \emph{scikit-learn} library \cite{scikit-learn}, where  
k-means is initialized with 
k-means++ \cite{Vassilvitskii}, 
and GMM is initialized with the output of k-means, making 
it more robust and preventing
it from breaking in high dimensions. 
The spectral clustering implementation of scikit-learn 
is based on \cite{Malik}. 
Kernel k-means is implemented according to Algorithm~\ref{kmeans_algo} while
kernel k-groups follows Algorithm~\ref{algo}.
Both will also be initialized with k-means++, unless specified otherwise.
We run every algorithms $5$ times with different initializations and then
we choose the result with the best objective function value. 
We evaluate the clustering quality by
the accuracy defined as
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi
\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij} ,
\end{equation}
where $\hat{Z}$ is the predicted label matrix, $Z$
is the ground truth label matrix, 
and $\pi$ is a permutation of $\{1,2,\dotsc,k\}$. 
Thus, the accuracy corresponds to the fraction
of correctly classified data points, and it is always between $[0,1]$.
For each setting, we show the average accuracy over $100$ Monte
Carlo trials, with error bars indicating the standard error.

@@

We first consider one-dimensional data for a two-class problem. We compare
kernel k-groups with k-means and GMM, as illustrated in 
Fig.~\ref{fig:1dexperiment}. The left panels show a mixture of Gaussians, and the right panels show a mixture of log Gaussians (see caption for details).
% have two mixtures of normal and lognormal distributions. 
Notice that in the kernel density estimation plots
for lognormal distribution, only kernel k-groups was able to distinguish
between the two classes. The accuracy results for both density estimation
cases are in Table~\ref{tb:dens}. We remark that kernel k-groups in
this one-dimensional example performed the same as the exact deterministic
Algorithm~\ref{algo1d} introduced in the Appendix.

\begin{table}[h]
\centering
\caption{
\label{tb:dens} Accuracy results for the density
estimation of Fig.~\ref{fig:1dexperiment}d--e.
}
\begin{tabular}{@{}l|ll@{}}
       Method        & normal & lognormal \\ \midrule[.5pt]
        kmeans       & 0.773  & 0.502     \\
         GMM         & \textbf{0.881}  & 0.513     \\
   kernel k-groups   & 0.802  & \textbf{0.847}  
\end{tabular}
\end{table}

@@

\begin{table}
\caption{
\label{tb:cigar_circle}
Clustering the data from Fig.~\ref{fig:2dscatter}.
For spcetral clustering, kernel k-means and kernel k-groups we use
the metric $\widetilde{\rho}_2$ (see \eqref{eq:rho_tilde}) for the data
in Fig.~\ref{fig:2dscatter}a, while $\widehat{\rho}_1$ (see \eqref{eq:rho_hat})
for the data in Fig.~\ref{fig:2dscatter}b. We have 800
points on each trial and 30 Monte Carlo runs for both datasets.
}
\centering
\begin{tabular}{@{}l|ll|ll@{}}
                   & \multicolumn{2}{c|}{Fig.~\ref{fig:2dscatter}a}  &
                     \multicolumn{2}{c}{Fig.~\ref{fig:2dscatter}b}  \\
       Method      & Accuracy & SEM     & Accuracy & SEM \\ \midrule[.5pt]
        kmeans       & 0.533 & 0.005  & 0.521   & 0.003 \\
         GMM         & 0.929 & 0.029  & 0.533   & 0.004 \\
 spectral clustering & 0.577 & 0.010  & 0.725   & 0.003 \\
    kernel k-means   & \textbf{1.000} & 0.000  & \textbf{1.000}   & 0.000 \\
   kernel k-groups   & \textbf{1.000} & 0.000  & \textbf{1.000}  & 0.000  
\end{tabular}
\end{table}

@@

\begin{table}
\caption{
\label{tb:dermatology_accuracy}
For the dataset \cite{Dua2017,Guvenir1998} (see also
Table~\ref{tb:dermatology}) we show the accuracy \eqref{eq:accuracy} and
the adjusted Rand index (aRand) of several methods.
In \cite{RizzoClustering} the authors 
obtained $\textnormal{aRand}=0.9195$ using an energy statistics
based method, while \cite{Kgroups} obtains $\textnormal{aRand}=0.9188$ 
where points with missing
entries are removed. Below we complete the missing entries with the mean.
If we remove the points with missing entries, kernel k-groups provides an
improvement of $\textnormal{accuracy}=0.9637$ and $\textnormal{aRand}=0.9396$.
}
\centering
\begin{tabular}{@{}l|ll@{}}
       Method      & Accuracy & aRand \\ \midrule[.5pt]
        kmeans       & 0.713 & 0.690 \\
         GMM         & 0.877 & 0.840 \\
 spectral clustering & 0.954 & 0.912 \\
    kernel k-means   & 0.751 & 0.851 \\
   kernel k-groups   & \textbf{0.962} & \textbf{0.936}
\end{tabular}
\end{table}
